---
title: "Homework 10 Example - Admixture, PCA, and DAPC"
author: "Nolan Helmstetter"
format:
  html:
    self-contained: true
editor: visual
---

# Viewing admixture results completed via the bash shell on tempest, using the tool admixture

```{r}
#| include: false
library(tidyverse)
library(dplyr)
library(sf)
library(ggplot2)
library(scatterpie)
library(adegenet)
library(vcfR)
```

Load admixture output

```{r}

# read sample names and extract
fam <- read_table("/home/g27n141/bioe-591-genomics/students/Helmstetter/hw_output/10/output/IDGS.int.fam", 
                  col_names = FALSE,
                  show_col_types = FALSE
)

```

Extract Individual IDs

```{r}
samples <- fam$X2   # individual IDs are usually column 2
```

Choose your K value (will be 2 for this demo)

```{r}
K <- 2
```

Read Q matrix: Ancestry proprtions for each individual

```{r}
q <- read_table("/home/g27n141/bioe-591-genomics/students/Helmstetter/hw_output/10/output/IDGS.int.2.Q",
                col_names = FALSE,
                show_col_types = FALSE
)

head(q)
```

Name the ancestry columns

```{r}
colnames(q) <- paste0("Cluster", 1:K)
```

Combine with sample names

```{r}
q_df <- q %>%
  mutate(sample = samples) %>%
  relocate(sample)
head(q_df)
```

Convert to long format for ggplot

```{r}
q_long <- q_df %>%
  pivot_longer(
    cols = starts_with("Cluster"),
    names_to = "cluster",
    values_to = "ancestry"
  ) %>%
  mutate(sample = factor(sample, levels = samples))
head(q_long)
```

```{r}
# plot
ggplot(q_long, aes(x = sample, y = ancestry, fill = cluster)) +
  geom_col(width = 1, color="white") +
  theme_bw() +
  labs(x = "Individual", y = "Ancestry proportion") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

Use regex functions in R to extract our cross-validation values. CV values are hidden in the .out files.

```{r}
cv_df <- tibble(file = "/home/g27n141/bioe-591-genomics/students/Helmstetter/hw_output/10/log/admix_step2-3840151.out") %>%
  mutate(
    text = map_chr(file, read_file),
    cv_line = str_extract(text, "CV error \\(K=\\d+\\):\\s*[-0-9.eE]+"),
    K  = str_match(cv_line, "CV error \\(K=(\\d+)\\)")[,2] |> as.integer(),
    CV = str_match(cv_line, ":\\s*([-0-9.eE]+)")[,2] |> as.numeric()
  ) %>%
  select(file, K, CV) %>%
  arrange(K)

```

# Plot samples and ancestry percentage in space

```{r}
# Load CSV File
csv <- read_csv("/home/g27n141/bioe-591-genomics/students/Helmstetter/hw_output/10/data/IDGS_BIOCLIM_Land_Soil_all_ed.csv", col_names = TRUE, show_col_types = FALSE
)

# Rename ID as sample
csv <- csv %>%
  rename(sample = ID)

```

Merge dfs for plotting

```{r}
q_df2 <- q_df %>%
  left_join(csv %>% select(sample, Lat, Long), by = "sample")

```

Make df into an sf object for plotting in space.

```{r}
q_sf <- st_as_sf(q_df2, coords = c("Long", "Lat"), crs = 4326)

```

Plot as pie chart showing ancestry (I just did this for some marten data) :)

```{r}
ggplot() +
  geom_scatterpie(
    data = q_df2,
    aes(x = Long, y = Lat),
    cols = c("Cluster1", "Cluster2"),
    pie_scale = 1.2,
    color = "black"
  ) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "none")


```

# Running a PCA

Load .vcf data
```{r}
vcf <- read.vcfR(
  file = "/home/g27n141/bioe-591-genomics/students/Helmstetter/hw_output/10/data/IDGS.recode.vcf",
  verbose = TRUE
)
```


Converts genotpyes in VCF and makes a DNbin object. A, C, G, T for each individual.
```{r}

dna <- vcfR2DNAbin(vcf, unphased_as_NA = F, consensus = T, extract.haps = F)
species_genind <- DNAbin2genind(dna)
species_genind


```

Deal with missing data and center genotypes (i.e., make laternative homozygotes equivalent, not translate to numierc values of 0 or 2 based on their allele count). Then run PCA on data.
```{r}

species_genind_scaled <- scaleGen(species_genind,NA.method="mean",scale=F)
species_pca <- prcomp(species_genind_scaled, center=F,scale=F)


```

Run Screeplot. While PC1 explains most of the variation in our data, variation appears to be fairly spread out. There is no sharp decline after PC1 and PC2. 
```{r}
screeplot(species_pca)

```
Plotting individuals in PC1 and PC2 space. Seperate by species. As expected, southern Idaho ground squirrels have more genetic variation than the smaller, isolated Northern Idaho ground squirrel populations. 
```{r}
pc <- data.frame(species_pca$x[,1:3])
pc$sample <- rownames(pc)

# Get speies from csv for each sample
pc <- pc %>%
  left_join(csv %>% select(sample, Species), by = "sample")

# Plot seperated by species
ggplot(pc, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 2) +
  theme_bw()+
  theme(panel.grid = element_blank())


```
# Running a Discriminant Analysis of Principal Components (DAPC)

Assign samples distinct, nonoverlapping clusters based on similarity in in PC space
```{r}
grp <- find.clusters(species_genind, n.pca = 50, n.clust = 2)
grp

```



Plot samples in PC1 and PC2 space seperating by defined clusters (above)
```{r}

pc$cluster <- grp$grp
ggplot(pc, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2) +
  geom_text(aes(label = sample), vjust = -0.5, size = 3) 


```

DAPC takes these hard-coded assignments and determines which principal components contribute to them, as well as any uncertainty in ancestry. (It is thus designed to discriminate on the basis of user-input populations, and should be used in combination with 
-means clustering or other approaches for detecting and / or designating structure.)


```{r}
# Keeps 50 principal components, Finds axes that maximize between-group differences for n.da = 2 groups
# na.da = 2 means we are asking for two discriminant axes, but since k = 2 groups, only 1 meaingful axis exists.
# We want to maximize variation between groups and minimize variation within groups
# Gives probabilities of each individual belonging to each group

dapc1 <- dapc(species_genind, pop = grp$grp, n.pca = 50, n.da = 2)
dapc1

```

The important part of this output are the posterior probabilities of ancestry assignments. We will extract these and turn it into a tibble with sample name, cluster, and ancestry probability:
Looks analgous to our admixture results.
```{r}
q <- as.data.frame(dapc1$posterior)
  q$sample <- rownames(q)
  q_long <- q |>
    pivot_longer(
      cols = -sample,
      names_to = "cluster",
      values_to = "ancestry"
    )
q_long

```

Plot ancestry estimates from DAPC
```{r}

ggplot(q_long, aes(x = sample, y = ancestry, fill = cluster)) +
  geom_col(width = 1, color = "white") +
  theme_bw() +
  labs(x = "Individual", y = "Assignment probability") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )


```

Compare results across differetn values of K
Uses BIC to evaluate results (i.e., model fit vs model complexity). This is not running a bayesian analysis despite using BIC.
```{r}

# Identify clusters
grp_auto <- find.clusters(
  species_genind,
  n.pca = 50,
  choose.n.clust = FALSE, 
  max.n.clust = 10,
  stat = "BIC"
)

# Print BIC values
grp_auto$Kstat


```

Plot BIC values for each K
Notice the elbow at K = 2, while BIC gets better as K increases, the biggest drop (or delta) is between 1 and 2. K = 2 appears to have the most meaningful jump in population structure.
```{r}
# plot!
plot(
  1:length(grp_auto$Kstat),
  grp_auto$Kstat,
  type = "b",
  xlab = "K",
  ylab = "BIC"
)

```
