---
title: "Homework 09: Kinship and Relatedness"
author: "Nolan Helmstetter"
format: html
editor: visual
---

Homework Assignment: Homework Assignment: comparing ngsRelate estimators of relatedness and kinship to those from `related::coancestry()`


Load R Environment and ngsRelate file.

```{r}
load("/home/g27n141/bioe-591-genomics/students/Helmstetter/hw_output/09/r_env.RData")
ngs<-read.table("/home/g27n141//bioe-591-genomics/students/Helmstetter/hw_output/09/example/Koala_MaxMissing10.recode.ngsrelate.out", header = TRUE)
```


Create matching IDs between data sets.

```{r}
relate <- kin_results$relatedness

# Get unique IDs for each pairwise comparison, ensuring you don't have duplicates (i.e., 1 & 2 and 2 & 1)
relate$pair <- paste(pmin(relate$ind1.id, relate$ind2.id),
                  pmax(relate$ind1.id, relate$ind2.id),
                  sep = "_")

ngs$pair <- paste(pmin(ngs$ida, ngs$idb),
                  pmax(ngs$ida, ngs$idb),
                  sep = "_")

# Merge df by pairwise comparison
merged <- merge(relate, ngs, by = "pair")

# Check for duplicates
any(duplicated(merged$pair))
```

Plot Several Pairwise Results Comparing relate to ngsRelate estimators. Here I will compare wang vs rab, quellergt vs rab, and dyadml vs theta using plots and Pearson's correlation coefficient.

```{r}
#| message: false
#| warning: false
library(ggplot2)
library(tidyr)
library(dplyr)
```

Calculate Pearson's correlation coefficient

```{r}
wang_rab<-cor(merged$wang, merged$rab, use = "complete.obs")
quell_rab<-cor(merged$quellergt, merged$rab, use = "complete.obs")
dyad_theta<-cor(merged$dyadml, merged$theta, use = "complete.obs")

```

Plot: Wang vs. rab

```{r}
ggplot() +
  geom_point(data = merged, aes(x = wang, y = rab), color = "black", alpha = 0.6) +
  annotate("text",
           x = min(merged$wang, na.rm = TRUE),
           y = max(merged$rab, na.rm = TRUE),
           hjust = 0,
           label = paste0("Pearson's Correlation Coefficient = ", round(wang_rab, 3))) +
  labs(x = "Wang", y = "rab") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black")
  )

```

Plot: quellergt vs. rab

```{r}
ggplot(merged, aes(x = quellergt, y = rab)) +
  geom_point(alpha = 0.6) +
  annotate("text",
           x = min(merged$quellergt, na.rm = TRUE),
           y = max(merged$rab, na.rm = TRUE),
           hjust = 0,
           label = paste0("Pearson's Correlation Coefficient = ", round(quell_rab, 3))) +
  labs(x = "quellergt", y = "rab") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black")
  )

```

Plot: dyadml vs. theta

```{r}
ggplot(merged, aes(x = dyadml, y = theta)) +
  geom_point(alpha = 0.6) +
  annotate("text",
           x = min(merged$dyadml, na.rm = TRUE),
           y = max(merged$theta, na.rm = TRUE),
           hjust = 0,
           label = paste0("Pearson's Correlation Coefficient = ", round(dyad_theta, 3))) +
  labs(x = "DyadML", y = "Theta") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = "black")
  )
```

